QueryMed: An Intuitive Federated SPARQL Query Builder for Biomedical RDF Data

### 2010-02-08 docs/www2010/paper.tex
fixing up paper formatting; removing irrelevant sections

% This is "www2010-sample.tex" copied from "www2005-sample.tex" V1.2 January 26 2004
% This file should be compiled with V1.4 of "www2010-submission.class"
%
% This example file demonstrates the use of the 'www2010-submission.cls'
% V1.4 LaTeX2e document class file. It is for those submitting
% articles to the WWW'04 Conference WHO DO NOT WISH TO 
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'www2010-submission.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V1.4) produces:
%       1) NO Permission Statement
%       2) WWW'04-specific conference (location) information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Julie Goetz (goetz@acm.org) or Adrienne Griscti (griscti@acm.org)
%
% Technical questions only to
% Gerald Murray (murray@acm.org)
% ===============================================================
%
% For tracking purposes - this is V1.2 - January 26 2004
\documentclass{www2010-submission}

@@

\maketitle
\begin{abstract}
We have developed an open-source SPARQL query builder and result set visualizer for biomedical data, QueryMed, that allows end users to easily construct and run translational medicine queries across multiple data sources. QueryMed is flexible enough to allow queries relevant to a wide range of biomedical topics, runs federated queries across multiple SPARQL endpoints, and is designed to be accessible to users who do not know the structure of the underlying ontologies or the SPARQL query language. The system allows users to select the data sources that they wish to use, drawing on their specialized domain knowledge to decide the most appropriate data sources to query.  Users can add additional data sources if they are interested in querying endpoints that are not in the default list. After retrieval of the initial result set, query results can be filtered to improve their relevance. As an advanced querying feature, the system also allows the user to exploit the underlying structure of the RDF data to improve query results. 
\end{abstract}

@@

RDF data in the semantic web, like the data in relational databases, is modeled in a structured format that can be queried using a standardized query language. The biomedical domain is among the early successes of the semantic web, due to the rapidity with which the biomedical community has made its data available in RDF triple stores. However, although a plethora of useful biomedical data is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and that also allow users to run federated queries on multiple SPARQL endpoints.

For example, a physician may know her patient’s personal information, symptoms, current medications, and genotype. She may wish to determine the patient’s treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and GO \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.  For example, to address her question, the physician might be interested in information on “coronary artery disease” (which she thinks is the cause of her patient’s symptoms). The information about the disease can be found in Diseasome. She might also want to know the set of drugs in DailyMed and Drugbank that can be used to treat coronary artery disease and that also will not react adversely with the patient’s current medications.  Information on clinical trials available for her patient can be found in LinkedCT.  Due to the large number of databases  that the physician needs to query in order to find an answer to her single question, she is likely to find useful a system that can automatically run queries over multiple data sources. Also, the physician may not know SPARQL query syntax, the location of the SPARQL endpoints, or the structure of the relevant ontologies. She is likely to want an intuitive way to query and to display the query result.  Developing intuitive ways to query multiple data sources and display results is both an important and a challenging problem.

@@

The semantic web can be viewed as global database system for the data available on the world wide web. Semantic web data is modeled by structured languages such as RDF and OWL, and can be queried using a standardized query language, SPARQL \cite{Berners-Lee}. The addition of structure to web data is intended to allow inferences to be automatically made by intelligent agents drawing on data from multiple sources.

RDF data is described by a directed, labeled graph, with relationships among data items represented  in triples containing a subject, an object and a predicate.  For example, we might want to express the concept, "Sam is a professor for 6.830." The corresponding triple expresses the relationship "is a professor for" between two entities, “Sam” and “6.830”. The relationship "is a professor for" is known as the predicate of the triple; "Sam" is the subject while "6.830" is the object.  In order for this statement to be machine readable, each of the entities in the statement ("Sam," "is a professor for," and "6.830") will be described by a Uniform Resource Identifier (URI) \cite{Manola}.  The Web Ontology Language (OWL) is a language that permits the definition of ontologies, allowing additional properties and constraints on relationships among data to be represented \cite{Smith}. Some parallels between relational databases and RDF data on the semantic web are illustrated in table 1 \cite{Berners-Lee}.

Although most web data are not currently available in semantic web formats, the biomedical community has been extremely proactive in making structured biomedical data available in RDF. Thus, the biomedical knowledge domain has been an early success of the semantic web \cite{Yip}. Many major biological and biomedical data resources, including Gene Ontology, DailyMed, LinkedCT, and Diseasome, are currently available as RDF triplestores.   Integrating biomedical data across multiple data sources and automatically extracting specific  knowledge from web resources are crucial tasks for physicians and biologists, and these semantic web resources represent valuable standardized repositories of information that can be automatically mined for biological knowledge.  The Semantic Web Health Care and Life Sciences Interest Group (HCLSIG) has been formed with the purpose of exploring the applications of the semantic web to the biomedical domain \cite{HCLSIG}.

@@

\subsection{SPARQL Query Language}

A SPARQL query performs a pattern-matching search on the underlying RDF graph of the desired resource.  Most SPARQL queries are constructed from basic graph patterns.  Like RDF triples, a basic graph pattern contains a subject, an object, and a predicate; however, in the basic graph pattern, the subject, object, or predicate may be replaced by variables.  The query result is the set of tuples matching the query pattern \cite{Prud'hommeaux}.

The SPARQL query language is designed to query semantic web data, just as SQL is designed to query relational data. However unlike SQL, SPARQL supports multiple query forms appropriate for  retrieving information from structured graph data. In SPARQL, the following types of queries are possible:
1. "ASK" query to test whether a graph contains  specific data of interest.  This type of query returns a boolean result.
2. "DESCRIBE" query to generate an RDF description of one or more resources.  This type of query returns a set of RDF triples.
3. "CONSTRUCT" query to create a custom RDF graph based on a query criteria. This can also be used to transform RDF data.  This type of query returns a graph matching the query criteria.
4. “SELECT” query to identify triples matching a query pattern.  This type of query returns a set of matching triples.

Our system supports a subset of queries of the “SELECT” type.

\subsection{The Query Federation Problem}

In the relational database world, data is contained within "data silos."  Schemas for different databases are likely to differ, even if the data that is stored is similar.  For example, suppose that two companies, Company A and Company B, construct independent databases,   Database 1 and Database 2 respectively, to describe all known drugs.  Company A may include “brand name,” “generic name,” “active compound,” “cost,” “indications,” and “protein target” information in its schema, while Company B may define the corresponding table in its database using a different schema.  In addition to discrepancies in table schemas, the data itself may be represented differently by the two companies. For example, the protein target of a drug to treat cardiovascular disease may be described as “ABCA1” in Database 1, and by a synonym, “ABC1,” in Database 2.  Therefore, it will be difficult to integrate the data between Database 1 and Database 2 without writing extensive application logic, even though the two databases contain highly overlapping information. In contrast, RDF data is represented in a form that is amenable for data integration spanning multiple data sources, because unique entities are described by unique Uniform Resource Identifiers (URIs).

However, because of the scale free nature of the web, locating the appropriate data endpoint to execute a specific query is a challenging problem. This is known as the query federation problem. While the SPARQL query language has some support for expressing queries that span multiple data sources, there is a need for improved systems able to orchestrate queries across multiple sources.

@@

QueryMed is implemented in Java in the backend and JavaScript in the frontend.  The JQuery library \cite{Resig} was used to develop an attractive user interface.  In the backend, the Jena library \cite{McBride} is used to run the SPARQL queries. QueryMed relies on the following remote resources:
SPARQL Endpoints:  QueryMed runs queries on multiple remote SPARQL endpoints that expose biomedical data.
Sources List: The default list of endpoints available to the user is stored in a source list on the proxy server.

@@

The user submits a query to the SPARQL endpoints via the proxy server.  The query is translated into a SPARQL query for each individual endpoint, and the results are returned to the user.  The specific components implemented in the proxy server are as follows.
Source Manager: The source manager reads the source list and populates the default query list on the user interface. It also keeps track of the default endpoints, the currently selected endpoints, and the endpoints that have been dynamically added.
Translator: This component is responsible for translating the user query into valid SPARQL syntax.
Run Query: The purpose of this service is to run SPARQL queries.  It takes input on source and keyword from the client, and returns  the query results as a JSON object. The structure of the JSON object is given in figure 4.
Get Properties: This service takes as input an individual source, and returns an array of all properties for that source.
Get Property Values: This service takes as input an individual source and a specific property.  It returns a list of the possible values that this property can take.
Advanced Queries: This service takes as input list of sources, properties, query terms, and logical operators, which are passed to the proxy server as a JSON object. The structure of this JSON object is given in figure 5. After constructing and executing the SPARQL query, it returns a set of query results as a JSON object in the format given in figure 4.

@@

The query interface is implemented in JavaScript, HTML and CSS. The main components of the user interface are as follows: 
Initial Query Interface: The user inputs their query  term into the provided text box (figure 6).
Add New Source: Allows the user to add a new data source to extend the query. (figure 8)
Dynamic Facet Creator/Filter: This component allows the user to select a  set of  data sources, load the properties available at these sources, and dynamically construct a complex SPARQL query connected by logical operators to take advantage of the structure of the data at each endpoint (figure 13).
Result Browser: The user can view their query results, organized by source. The results are presented in a table with pagination.  The user can choose the number of results viewed at a time, search the columns based on some text value and also sort the columns (figure 14).

@@

\begin{verbatim}
SELECT ?projection_1 ?projection_2 ... ?projection_n
WHERE {?x source:property ? projection 
	FILTER regex(? projection, '" + input +"', 'i)
}
... //Other property filers are to follow
\end{verbatim}
Minimal SPARQL query structure in the “Query All” case

@@

\begin{verbatim}
{
     "bindings":
     [
     	{"source" : "source_label",
     	 "uri" : "http://source.com/sparql.",
     	 "vars" : ["variable1","variable2"],
     	 "count":2,
     	 "results" : 
     		[
     		 {"variable1": "Value 1", 		   "variable1": "Value 2"},
     		 {"variable1": "Another Value 1", 		   "variable1": "Another Value 2"},

     		]
     	}
     	,
	{ //Another Source and Result Description
	...
	},
	...
     ]
 }
\end{verbatim}
JSON returned from the RunQuery service

\begin{verbatim}
{
   "http://some-source/sparql": {
                            "property1": {"value1": "AND"},
	              "property2": {"value 2": "OR"},
	              "property2": {"value 3": "FILTER"}
	           }
	}
   //Other sources ...
}
\end{verbatim}
JSON constructed to send the structure of a SPARQL query for an advanced query with logical operators

@@

\begin{verbatim}
SELECT DISTINCT ?property
WHERE { [] ?property [] }
ORDER BY ?property
\end{verbatim}

@@

\begin{verbatim}
SELECT distinct ?disease WHERE { 
{?x <http://www.w3.org/2000/01/rdf-schema#label> 
?disease 
FILTER regex(?disease, "coronary artery disease", "i"). 
?x <http://www4.wiwiss.fu-berlin.de/diseasome/resource/diseasome/class> 
<http://www4.wiwiss.fu-berlin.de/diseasome/resource/diseaseClass/Cardiovascular>} 
UNION { ?x <http://www4.wiwiss.fu-berlin.de/diseasome/resource/diseasome/associatedGene>
 <http://www4.wiwiss.fu-berlin.de/diseasome/resource/genes/ABCA1>.}
}
\end{verbatim}
A complex SPARQL query that takes advantage of the underlying data structure can be dynamically constructed using the advanced query feature of the QueryMed system.

@@

The source code for the Querymed system is available at the QueryMed Google Code project: 
http://code.google.com/p/querymed/

A video displaying a sample use case can be found at: 
http://www.youtube.com/watch?v=JCr7vryS9Vg

>>>>

% This is "www2010-sample.tex" copied from "www2005-sample.tex" V1.2 January 26 2004
% This file should be compiled with V1.4 of "www2010-submission.class"
%
% This example file demonstrates the use of the 'www2010-submission.cls'
% V1.4 LaTeX2e document class file. It is for those submitting
% articles to the WWW'04 Conference WHO DO NOT WISH TO 
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'www2010-submission.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V1.4) produces:
%       1) NO Permission Statement
%       2) WWW'04-specific conference (location) information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Julie Goetz (goetz@acm.org) or Adrienne Griscti (griscti@acm.org)
%
% Technical questions only to
% Gerald Murray (murray@acm.org)
% ===============================================================
%
% For tracking purposes - this is V1.2 - January 26 2004
\documentclass{www2010-submission}
\usepackage{moreverb}

@@

\maketitle
\begin{abstract}
We have developed an open-source SPARQL query builder and result set visualizer for biomedical data, QueryMed, that allows end users to easily construct and run translational medicine queries across multiple data sources. QueryMed is flexible enough to allow queries relevant to a wide range of biomedical topics, runs federated queries across multiple SPARQL endpoints, and is designed to be accessible to users who do not know the structure of the underlying ontologies or the SPARQL query language. The system allows users to select the data sources that they wish to use, drawing on their specialized domain knowledge to decide the most appropriate data sources to query.  Users can add additional data sources if they are interested in querying endpoints that are not in the default list. After retrieval of the initial result set, query results can be filtered to improve their relevance. As an advanced search feature, the system also allows the user to exploit the underlying structure of the RDF data to improve query results. 
\end{abstract}

@@

The biomedical domain is among the early successes of the semantic web, due to the rapidity with which the biomedical community has made its data available in RDF triple stores. However, although a plethora of useful biomedical data is currently available in RDF, there is a need for easy-to use systems that do not require the end user to have knowledge of the underlying  structure of the data, and that also allow users to run federated queries on multiple SPARQL endpoints.

For example, a physician may know her patient's personal information, symptoms, current medications, and genotype. She may wish to determine the patient's treatment plan and identify clinical trials for which the patient is eligible.  Although the physician has a single question--based on the information I have about this patient, what is the best treatment plan and set of clinical trials available?--there is no single data source that the physician can use to answer this question. The information that the physician needs must be gathered from numerous data sources such as Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome,  and GO \cite{Pubmed, DailyMed, Drugbank, LinkedCT, Diseasome, GO}. Her question must be broken up into discrete pieces that can be executed individually at one data source at a time.  For example, to address her question, the physician might be interested in information on “coronary artery disease” (which she thinks is the cause of her patient’s symptoms). The information about the disease can be found in Diseasome. She might also want to know the set of drugs in DailyMed and Drugbank that can be used to treat coronary artery disease.  Information on clinical trials available for her patient can be found in LinkedCT.  Due to the large number of databases  that the physician needs to query in order to find an answer to her single question, she is likely to find useful a system that can automatically run queries over multiple data sources. Also, the physician may not know SPARQL query syntax, the location of the SPARQL endpoints, or the structure of the relevant ontologies. She is likely to want an intuitive way to query and to display the query result.  Developing intuitive ways to query multiple data sources and display results is both an important and a challenging problem.

@@

Although most web data are not currently available in semantic web formats, the biomedical knowledge domain has been an early success of the semantic web \cite{Yip}. Many major biological and biomedical data resources, including Gene Ontology, DailyMed, LinkedCT, and Diseasome, are currently available as RDF triplestores.   Integrating biomedical data across multiple data sources and automatically extracting specific  knowledge from web resources are crucial tasks for physicians and biologists, and these semantic web resources represent valuable standardized repositories of information that can be automatically mined for biological knowledge.  The Semantic Web Health Care and Life Sciences Interest Group (HCLSIG) has been formed with the purpose of exploring the applications of the semantic web to the biomedical domain \cite{HCLSIG}.

@@

@@

QueryMed is implemented in Java in the backend and JavaScript in the frontend.  The JQuery library \cite{Resig} was used to develop an attractive user interface.  In the backend, the Jena library \cite{McBride} is used to run the SPARQL queries. QueryMed relies on the following remote resources:
\begin{itemize}
\item \textbf{SPARQL Endpoints}:  QueryMed runs queries on multiple remote SPARQL endpoints that expose biomedical data.
\item \textbf{Sources List}: The default list of endpoints available to the user is stored in a source list on the proxy server.
\end{itemize}

@@

The user submits a query to the SPARQL endpoints via the proxy server.  The query is translated into a SPARQL query for each individual endpoint, and the results are returned to the user.  The specific components implemented in the proxy server are as follows.
\begin{itemize}
\item \textbf{Source Manager:} The source manager reads the source list and populates the default query list on the user interface. It also keeps track of the default endpoints, the currently selected endpoints, and the endpoints that have been dynamically added.
\item \textbf{Translator:} This component is responsible for translating the user query into valid SPARQL syntax.
\item \textbf{Run Query:} The purpose of this service is to run SPARQL queries.  It takes input on source and keyword from the client, and returns  the query results as a JSON object. The structure of the JSON object is given in figure 4.
\item \textbf{Get Properties:} This service takes as input an individual source, and returns an array of all properties for that source.
\item \textbf{Get Property Values:} This service takes as input an individual source and a specific property.  It returns a list of the possible values that this property can take.
\item \textbf{Advanced Queries:} This service takes as input list of sources, properties, query terms, and logical operators, which are passed to the proxy server as a JSON object. The structure of this JSON object is given in figure 5. After constructing and executing the SPARQL query, it returns a set of query results as a JSON object in the format given in figure 4.
\end{itemize}

@@

The query interface is implemented in JavaScript, HTML and CSS. The main components of the user interface are as follows: 
\begin{itemize}
\item \textbf{Initial Query Interface:} The user inputs their query  term into the provided text box (figure 6).
\item \textbf{Add New Source:} Allows the user to add a new data source to extend the query. (figure 8)
\item \textbf{Dynamic Facet Creator/Filter:} This component allows the user to select a  set of  data sources, load the properties available at these sources, and dynamically construct a complex SPARQL query connected by logical operators to take advantage of the structure of the data at each endpoint (figure 13).
\item \textbf{Result Browser:} The user can view their query results, organized by source. The results are presented in a table with pagination.  The user can choose the number of results viewed at a time, search the columns based on some text value and also sort the columns (figure 14).
\end{itemize}

@@

\begin{verbatimtab}
SELECT ?projection_1 ?projection_2 ... 
	?projection_n
WHERE {?x source:property ? projection 
	FILTER regex(? projection, '" + 
	input +"', 'i)
}
... //Other property filers are to follow
\end{verbatimtab}
Minimal SPARQL query structure in the “Query All” case

@@

\begin{verbatimtab}
{
     "bindings":
     [
     	{"source" : "source_label",
     	 "uri" : "http://source.com/sparql.",
     	 "vars" : ["variable1","variable2"],
     	 "count":2,
     	 "results" : 
     		[
     		 {"variable1": "Value 1",
		  "variable1": "Value 2"},
     		 {"variable1": "Another Value 1",
		  "variable1": "Another Value 2"},

     		]
     	}
     	,
	{ //Another Source and Result Description
	...
	},
	...
     ]
 }
\end{verbatimtab}
JSON returned from the RunQuery service

\begin{verbatimtab}
{
   "http://some-source/sparql": {
   	"property1": {"value1": "AND"},
	"property2": {"value 2": "OR"},
	"property2": {"value 3": "FILTER"}
	           }
	}
   //Other sources ...
}
\end{verbatimtab}
JSON constructed to send the structure of a SPARQL query for an advanced query with logical operators

@@

\begin{verbatimtab}
SELECT DISTINCT ?property
WHERE { [] ?property [] }
ORDER BY ?property
\end{verbatimtab}

@@

\begin{verbatimtab}
SELECT distinct ?disease WHERE { 
{?x <http://www.w3.org/2000/01/rdf-schema#label> 
?disease 
FILTER regex(?disease, 
"coronary artery disease", "i"). 
?x <http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseasome/class> 
<http://www4.wiwiss.fu-berlin.de/diseasome/
resource/diseaseClass/Cardiovascular>} 
UNION { ?x <http://www4.wiwiss.fu-berlin.de/
diseasome/resource/diseasome/associatedGene>
 <http://www4.wiwiss.fu-berlin.de/diseasome/
 resource/genes/ABCA1>.}
}
\end{verbatimtab}
A complex SPARQL query that takes advantage of the underlying data structure can be dynamically constructed using the advanced query feature of the QueryMed system.

@@

The source code for the Querymed system is available at the QueryMed Google Code project: \\
http://code.google.com/p/querymed/

A video displaying a sample use case can be found at: \\
http://www.youtube.com/watch?v=JCr7vryS9Vg
